---
title: "Script 00 — Data Preprocessing & Variable Engineering"
subtitle: "TFM: Gender Wage Gap in Spain | MESIO UPC-UB | Julia Anglada Lomaeva | 2025-26"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 5.5, dpi = 150)
```

##00_preprocessing.R — Por qué cada decisión

### Tablas de decodificación: El INE entrega los microdatos con códigos numéricos/alfanuméricos (SEXO=1 para hombre, SEXO=6 para mujer). El 6 para mujer es un código poco intuitivo que causa errores silenciosos si no se documenta. Cada left_join mapea un código INE a una etiqueta interpretable en inglés. Los left_join en lugar de mutate/case_when son más legibles y más fáciles de verificar que las etiquetas sean correctas.

### Type harmonization: Todos los códigos INE se convierten a character antes del join. Si los códigos son numéricos en el raw data y character en la tabla de decodificación, el left_join falla silenciosamente (sin error, pero todas las filas quedan NA). Es uno de los bugs más difíciles de detectar. El código lo documenta explícitamente con el comentario # Silent type mismatches cause rows to silently drop.

### Filtrado de observaciones: Se eliminan salarios por hora < €2 (aproximadamente 0.3% de la muestra). El SMI de 2022 era €7.47/hr pero se usa €2 como corte conservador porque algunos contratos legítimos (becarios, contratos de aprendizaje) pueden estar por debajo del SMI. El filtro de €400 anuales elimina contratos de muy corta duración (días en el mes de referencia de octubre) que distorsionarían el hourly wage al dividir por horas anuales.

### Factor encoding: El factor gender se define con levels = c("Male", "Female") para que Male sea la referencia. En R, el primer nivel de un factor es la referencia en regresión. Esto es estándar en la literatura de wage gap: el coeficiente de genderFemale representa la penalización salarial femenina relativa al hombre.

### Listwise deletion: Se aplica solo a las variables del modelo principal. Las variables de robustness check (nationality, solo para DML) no entran en la eliminación principal para preservar el mayor n posible para OLS y OB.

```{r main}
# =============================================================================
# TFM — Gender Wage Gap in Spain
# MESIO UPC-UB | Julia Anglada Lomaeva | 2025-26
# -----------------------------------------------------------------------------
# SCRIPT 01: Variable Selection, Decoding & Preprocessing
# -----------------------------------------------------------------------------
# This script does three things in order:
#   1. Decodes ALL relevant INE codes into interpretable English labels
#   2. Engineers the dependent variable (log hourly wage)
#   3. Builds THREE nested model specifications:
#        Spec 1 → OLS Baseline (Mincer core, 12 controls)
#        Spec 2 → OLS Extended + Oaxaca-Blinder (adds MERCADO + REGULACION)
#        Spec 3 → DML (all of Spec 2 + TIPOPAIS nationality control)
#
# Variable selection rationale is documented inline with academic references.
# Excluded variables are listed explicitly with reasons — the tribunal will ask.
# =============================================================================

library(tidyverse)

# Load raw EES 2022 microdata
# Source: INE, Encuesta de Estructura Salarial 2022
# https://www.ine.es/dyngs/INEbase/es/operacion.htm?c=Estadistica_C&cid=1254736177025
load("~/Downloads/datos_2022/R/EES_2022.RData")
df_raw <- Microdatos

head(df_raw)
summary(df_raw)

cat("Raw dataset loaded:", nrow(df_raw), "observations,", ncol(df_raw), "variables\n")
```

```{r}
# =============================================================================
# SECTION 1: DECODING TABLES
# =============================================================================
# The INE releases EES microdata with numeric/alphanumeric codes. Every join
# below maps those codes to interpretable English labels following the official
# INE design record (dr_EES_2022.xlsx, sheets "Diseño", "Tablas1", "Tablas2").
# All mapping tables are built from the official INE code dictionaries.
# =============================================================================

# ── 1.1 Gender (SEXO → TSexo) ──────────────────────────────────────────────
# Codes: 1 = Hombre, 6 = Mujer
# Note: INE uses 1/6 (not 1/2) — a common source of silent merge failures
# Reference level: Male (standard in wage gap literature, Blau & Kahn 2017)
appendix_gender <- tibble(
  INE_Code = c("1", "6"),
  gender   = c("Male", "Female")
)

# ── 1.2 Education (ESTU → TTitula) ─────────────────────────────────────────
# 7 ordered levels from ISCED classification
# Ordered factor: tells OLS and RF that Bachelor > Primary (not just different)
# Human capital theory: Mincer (1974), Goldin (2014)
appendix_education <- tibble(
  INE_Code  = as.character(1:7),
  education = c(
    "Less than Primary",        # ISCED 0
    "Primary Education",        # ISCED 1
    "Lower Secondary",          # ISCED 2
    "Upper Secondary",          # ISCED 3
    "Vocational Training",      # ISCED 4/5 — FP Grado Superior
    "University Diploma",       # ISCED 6 — Diplomatura (3-year degree)
    "Bachelors/Masters/PhD"     # ISCED 7 — Grado/Máster/Doctorado
  )
)

# ── 1.3 Working Day Type (TIPOJOR → TJornad) ───────────────────────────────
# 1 = Full-time, 2 = Part-time
# IMPORTANT: we build log_hourly_wage (= log(RETRINOIN / JAP)) as the
# dependent variable, which already controls for hours in a continuous,
# precise way. TIPOJOR is then a label for contract modality, not hours.
# Including both log_hourly_wage AND TIPOJOR is fine — they capture different
# things (hourly price vs. contract type). See Goldin (2014) on part-time
# penalties operating beyond the pure hours effect.
appendix_working_day <- tibble(
  INE_Code    = c("1", "2"),
  working_day = c("Full-time", "Part-time")
)

# ── 1.4 Contract Type (TIPOCON → TContra) ──────────────────────────────────
# 1 = Indefinite (permanent), 2 = Fixed-term (temporary)
# Temporary contracts are associated with lower wages and higher female
# incidence in Spain. Dolado et al. (2002) document the "dual labour market"
# penalty in the Spanish context — one of the largest in the OECD.
appendix_contract_type <- tibble(
  INE_Code      = c("1", "2"),
  contract_type = c("Permanent", "Temporary")
)

# ── 1.5 Sector (CNACE → TCNAE) ─────────────────────────────────────────────
# 27 CNAE-2009 aggregated groups (B0 through S0)
# Horizontal occupational segregation by sector is one of the two main
# mechanisms explaining the Spanish gender wage gap (De la Rica et al. 2008).
# These groups follow the official EES aggregation, not full CNAE 2-digit.
appendix_sector <- tibble(
  INE_Code = c("B0","C1","C2","C3","C4","C5","C6","C7","C8",
               "D0","E0","F0","G1","G2","H1","H2","I0","J0",
               "K0","L0","M0","N0","O0","P0","Q0","R0","S0"),
  sector   = c(
    "Mining and Quarrying",
    "Manufacturing: Food, Textiles and Leather",
    "Manufacturing: Wood and Paper",
    "Manufacturing: Printing",
    "Manufacturing: Chemical and Pharma",
    "Manufacturing: Non-metallic Mineral Products",
    "Manufacturing: Metallurgy and Metal Products",
    "Manufacturing: IT, Electronics and Machinery",
    "Manufacturing: Vehicles and Furniture",
    "Electricity, Gas and Steam Supply",
    "Water Supply and Waste Management",
    "Construction",
    "Wholesale Trade and Motor Vehicles",
    "Retail Trade",
    "Transport (Land, Water, Air)",
    "Warehousing and Postal Activities",
    "Hospitality (Accommodation and Food)",
    "Information and Communications (ICT)",
    "Financial and Insurance Activities",
    "Real Estate Activities",
    "Professional, Scientific and Technical Activities",
    "Administrative and Support Services",
    "Public Administration and Social Security",
    "Education",
    "Human Health and Social Work Activities",
    "Arts, Entertainment and Recreation",
    "Other Service Activities"
  )
)

# ── 1.6 Occupation (CNO1 → TCNO) ───────────────────────────────────────────
# 17 CNO-2011 major groups (A0 through Q0)
# Vertical segregation by occupation is the second key mechanism in the
# Spanish wage gap literature (De la Rica et al. 2008, Muñoz-Bullón 2010).
# The glass ceiling manifests precisely in the underrepresentation of women
# at the top of this classification (group A0: Managers and Directors).
appendix_occupation <- tibble(
  INE_Code   = c("A0","B0","C0","D0","E0","F0","G0","H0","I0",
                 "J0","K0","L0","M0","N0","O0","P0","Q0"),
  occupation = c(
    "Managers and Directors",
    "Health and Education Professionals",
    "Science and Engineering Professionals",
    "Associate Professionals",
    "Office Clerks (Back Office)",
    "Customer Service Clerks",
    "Sales and Catering Workers",
    "Health and Care Services",
    "Protective and Security Services",
    "Skilled Agri, Forest and Fishery",
    "Skilled Construction Workers",
    "Skilled Manufacturing Workers",
    "Stationary Plant and Machine Operators",
    "Mobile Plant Operators and Drivers",
    "Elementary Service Occupations",
    "Elementary Laborers (Agri/Const/Manuf)",
    "Armed Forces"
  )
)

# ── 1.7 Region (NUTS1 → TNUTS) ─────────────────────────────────────────────
# 7 NUTS-1 macro-regions
# Spain's regional labour markets differ substantially in industrial
# composition, unionisation rates and wage levels (Simón et al. 2008).
# The NUTS-1 aggregation is a reasonable compromise between regional
# granularity and sample size per cell in the regression.
appendix_region <- tibble(
  INE_Code = as.character(1:7),
  region   = c(
    "Northwest",        # Galicia, Asturias, Cantabria
    "Northeast",        # País Vasco, Navarra, La Rioja, Aragón
    "Madrid",           # Comunidad de Madrid
    "Center",           # Castilla y León, Castilla-La Mancha, Extremadura
    "East",             # Cataluña, Valencia, Baleares
    "South",            # Andalucía, Murcia, Ceuta, Melilla
    "Canary Islands"    # Canarias
  )
)

# ── 1.8 Age Group (ANOS2 → TEdad) ──────────────────────────────────────────
# 6 age cohorts — ordered factor (younger < older)
# Age proxies general labour market experience when firm tenure is also
# included, following the two-component experience decomposition in
# Topel (1991): total experience = general + firm-specific.
appendix_age <- tibble(
  INE_Code  = c("01","02","03","04","05","06"),
  age_group = c("<19 years", "20-29", "30-39", "40-49", "50-59", ">59 years")
)

# ── 1.9 Firm Size (ESTRATO2 → TTrabaj) ─────────────────────────────────────
# 3 strata: Small (1-49), Medium (50-199), Large (≥200)
# Larger firms pay more on average (size-wage premium, Oi & Idson 1999) and
# have more formalised HR practices that may reduce discriminatory pay.
# Muñoz-Bullón & Sánchez-Bueno (2011) show firm size is a significant
# predictor of the gender gap in Spain.
appendix_firm_size <- tibble(
  INE_Code  = c("1", "2", "3"),
  firm_size = c("Small (1-49)", "Medium (50-199)", "Large (>200)")
)

# ── 1.10 Firm Sector / Ownership (CONTROL → TContro) ───────────────────────
# 1 = Public, 2 = Private
# The gender pay gap is systematically smaller in the public sector due to
# transparent pay scales and formalised grading systems (Christofides &
# Pashardes 2002). Including this variable is standard in Spanish wage gap
# studies (Palencia et al. 2014).
appendix_firm_sector <- tibble(
  INE_Code    = c("1", "2"),
  firm_sector = c("Public", "Private")
)

# ── 1.11 Supervisory Role (RESPONSA → T1Sino) ──────────────────────────────
# 0 = No supervisory responsibility, 1 = Yes
# Controls for within-occupation hierarchy — the glass ceiling operates
# precisely here. Without this control, part of the glass ceiling effect
# would be absorbed into the occupation fixed effect, understating the
# direct gender penalty at the top of hierarchies.
appendix_supervisory <- tibble(
  INE_Code         = c("1", "0"),
  supervisory_role = c("Yes", "No")
)

# ── 1.12 NEW: Market Scope (MERCADO → TMercad) ─────────────────────────────
# 1=Local/Regional, 2=National, 3=EU, 4=Global
# Exporting firms pay wage premiums (exporter wage premium, Bernard &
# Jensen 1999), and these premiums are distributed unequally by gender:
# Zweimüller et al. (2008) and Blau & Kahn (2017) show that the gender gap
# is larger in internationally competitive firms, where performance-related
# pay — which disadvantages women on average — plays a greater role.
# This is one of the EES-specific variables not available in most datasets.
appendix_market <- tibble(
  INE_Code     = c("1", "2", "3", "4"),
  market_scope = c("Local/Regional", "National", "European Union", "Global")
)

# ── 1.13 NEW: Collective Bargaining Level (REGULACION → TRegula) ────────────
# 1=National sector, 2=Regional/provincial sector, 3=Company, 4=Workplace, 5=Other
# Card, Cardoso & Kline (2016) show that collective bargaining structure
# explains 20-30% of between-firm wage variance. In Spain specifically,
# company-level agreements (codes 3-4) give firms substantially more
# discretion to set discretionary pay supplements — which is precisely where
# gender discrimination operates (Izquierdo et al. 2003, Bentolila &
# Saint-Paul 1992). We collapse into a binary: centralised (1-2) vs.
# decentralised (3-5), following the approach in Simón et al. (2008).
appendix_bargaining <- tibble(
  INE_Code             = as.character(1:5),
  collective_bargaining = c(
    "Centralised",      # 1: National/sectorial estatal
    "Centralised",      # 2: Regional/provincial sector
    "Decentralised",    # 3: Company or group agreement
    "Decentralised",    # 4: Workplace agreement
    "Decentralised"     # 5: Other (typically firm-specific)
  )
)

# ── 1.14 NEW: Nationality (TIPOPAIS → TPais) ───────────────────────────────
# 1 = Spain, 2 = Rest of world
# Used ONLY in Spec 3 (DML robustness). Immigrant workers have systematically
# different wage profiles for reasons unrelated to gender discrimination
# (selection into migration, language barriers, credential recognition).
# Including it prevents the gender estimator from picking up immigrant-female
# overlap effects. NOT included in OLS main specifications because the binary
# variable is too coarse to capture immigrant heterogeneity — it would absorb
# variance without adding clean identification (Amuedo-Dorantes & De la Rica 2007).
appendix_nationality <- tibble(
  INE_Code    = c("1", "2"),
  nationality = c("Spanish", "Foreign")
)
```

```{r}
# =============================================================================
# SECTION 2: TYPE HARMONIZATION
# =============================================================================
# All INE codes in the microdata must be character before joining.
# Silent type mismatches (numeric vs. character) cause rows to silently drop
# from left_joins — the worst kind of bug because the dataset looks fine.
# =============================================================================

df_typed <- df_raw %>%
  mutate(across(
    c(SEXO, ESTU, TIPOJOR, TIPOCON, CNACE, CNO1,
      NUTS1, ANOS2, ESTRATO2, CONTROL, RESPONSA,
      MERCADO, REGULACION, TIPOPAIS),
    as.character
  ))

cat("Type harmonization complete.\n")


# =============================================================================
# SECTION 3: JOIN ALL MAPPING TABLES
# =============================================================================

df_decoded <- df_typed %>%
  left_join(appendix_gender,     by = c("SEXO"      = "INE_Code")) %>%
  left_join(appendix_education,  by = c("ESTU"      = "INE_Code")) %>%
  left_join(appendix_working_day,by = c("TIPOJOR"   = "INE_Code")) %>%
  left_join(appendix_contract_type, by = c("TIPOCON" = "INE_Code")) %>%
  left_join(appendix_sector,     by = c("CNACE"     = "INE_Code")) %>%
  left_join(appendix_occupation, by = c("CNO1"      = "INE_Code")) %>%
  left_join(appendix_region,     by = c("NUTS1"     = "INE_Code")) %>%
  left_join(appendix_age,        by = c("ANOS2"     = "INE_Code")) %>%
  left_join(appendix_firm_size,  by = c("ESTRATO2"  = "INE_Code")) %>%
  left_join(appendix_firm_sector,by = c("CONTROL"   = "INE_Code")) %>%
  left_join(appendix_supervisory,by = c("RESPONSA"  = "INE_Code")) %>%
  left_join(appendix_market,     by = c("MERCADO"   = "INE_Code")) %>%
  left_join(appendix_bargaining, by = c("REGULACION"= "INE_Code")) %>%
  left_join(appendix_nationality,by = c("TIPOPAIS"  = "INE_Code"))

# Verify join quality — any unmatched codes appear as NAs
cat("\nJoin quality check (NAs per key variable):\n")
key_vars <- c("gender","education","working_day","contract_type","sector",
              "occupation","region","age_group","firm_size","firm_sector",
              "supervisory_role","market_scope","collective_bargaining","nationality")
for (v in key_vars) {
  n_na <- sum(is.na(df_decoded[[v]]))
  pct  <- round(n_na / nrow(df_decoded) * 100, 2)
  cat(sprintf("  %-25s: %6d NAs (%s%%)\n", v, n_na, pct))
}


# =============================================================================
# SECTION 4: DEPENDENT VARIABLE ENGINEERING
# =============================================================================

df_engineered <- df_decoded %>%
  mutate(
    # Gross annual salary (non-IT related) — the standard wage measure in EES
    # RETRINOIN: "Salario Bruto Anual No Derivado de IT"
    # We exclude RETRIIN (IT-derived salary) because incapacity spells
    # introduce a selection mechanism unrelated to structural wage-setting.
    annual_salary = as.numeric(RETRINOIN),

    # Annual pacted working hours — used to construct the hourly wage
    # JAP: "Jornada Anual Pactada"
    annual_hours = as.numeric(JAP),

    # Hourly wage: removes the part-time confound
    # Without this division, total annual earnings conflate hourly price
    # with hours worked — and women work part-time at ~4x the rate of men
    # in Spain (Eurostat LFS 2022), creating a mechanical downward bias
    # in any gender coefficient on annual wages.
    hourly_wage = annual_salary / annual_hours,

    # Log transformation: standard in wage regression since Mincer (1974)
    # Rationale: (1) approximates log-normality of wages; (2) stabilises
    # variance; (3) renders coefficients interpretable as % differentials
    # via exp(beta) - 1.
    log_hourly_wage = log(hourly_wage),

    # Firm tenure in years — the key human capital variable for firm-specific
    # experience (Topel 1991). ANOANTI is already numeric; we just rename.
    tenure = as.numeric(ANOANTI),

    # Tenure squared — captures the concave returns to experience
    # (rising then flattening wage-tenure profile). Without this term,
    # OLS linearly extrapolates experience returns, which is empirically
    # wrong and introduces systematic residual patterns.
    tenure_sq = tenure^2,

    # Overtime hours — used for descriptive purposes only, NOT in models
    # (it is partly endogenous to sector and occupation)
    overtime_hours = as.numeric(HEXTRA)
  )
```

```{r}
# =============================================================================
# SECTION 5: FILTERING
# =============================================================================
# Two principled exclusion criteria. Both are documented so the tribunal can
# see exactly what fraction of observations is affected.

n_before <- nrow(df_engineered)

df_filtered <- df_engineered %>%
  filter(
    # Exclude implausible annual salaries below €400
    # Rationale: these represent data entry errors, very short employment
    # spells (a few days in October with partial salary), or administrative
    # records that do not reflect a wage relationship. €400 corresponds to
    # roughly one week at minimum wage — a conservative lower bound.
    annual_salary > 400,

    # Exclude implausible hourly wages below €2
    # Rationale: the 2022 Spanish minimum wage (SMI) was €7.47/hour for
    # full-time workers. Values below €2 indicate either JAP coding errors
    # or the same short-spell issue above. We use €2 rather than SMI to
    # be conservative — some legitimate cases (traineeships, special
    # contracts for young workers) may fall between €2 and €7.47.
    hourly_wage > 2,

    # Exclude zero or negative hourly wages (log undefined)
    hourly_wage > 0,

    # Exclude missing annual hours (cannot construct hourly wage)
    !is.na(annual_hours),
    annual_hours > 0
  )

n_after  <- nrow(df_filtered)
n_removed <- n_before - n_after
cat(sprintf("\nFiltering: removed %d observations (%.2f%% of sample)\n",
    n_removed, n_removed / n_before * 100))
cat(sprintf("Analytical sample: %d observations\n", n_after))
```

```{r}
# =============================================================================
# SECTION 6: FACTOR ENCODING
# =============================================================================
# Ordered factors: education and age_group — their ordinal structure is
# meaningful and should be communicated to tree-based models (RF, DML).
# OLS with ordered factors uses polynomial contrasts by default in R;
# consider using unordered + reference level for cleaner OLS output.

df_factors <- df_filtered %>%
  mutate(

    # ── Treatment variable ────────────────────────────────────────────────
    # Male as reference: standard in the wage gap literature
    gender = factor(gender, levels = c("Male", "Female")),

    # ── ORDERED factors (meaningful cardinal ranking) ─────────────────────
    education = factor(education,
      ordered = TRUE,
      levels  = c("Less than Primary", "Primary Education", "Lower Secondary",
                  "Upper Secondary", "Vocational Training",
                  "University Diploma", "Bachelors/Masters/PhD")),

    age_group = factor(age_group,
      ordered = TRUE,
      levels  = c("<19 years", "20-29", "30-39", "40-49", "50-59", ">59 years")),

    firm_size = factor(firm_size,
      ordered = TRUE,
      levels  = c("Small (1-49)", "Medium (50-199)", "Large (>200)")),

    market_scope = factor(market_scope,
      ordered = TRUE,
      levels  = c("Local/Regional", "National", "European Union", "Global")),

    # ── UNORDERED factors (nominal categories) ────────────────────────────
    working_day          = factor(working_day),
    contract_type        = factor(contract_type),
    sector               = factor(sector),
    occupation           = factor(occupation),
    region               = factor(region),
    firm_sector          = factor(firm_sector),
    supervisory_role     = factor(supervisory_role),
    collective_bargaining= factor(collective_bargaining,
                                  levels = c("Centralised", "Decentralised")),
    nationality          = factor(nationality, levels = c("Spanish", "Foreign"))
  )

cat("\nFactor encoding complete.\n")
cat("Summary of treatment variable:\n")
print(table(df_factors$gender, useNA = "always"))
```

```{r}
# =============================================================================
# SECTION 7: LISTWISE DELETION — DROP NAs IN MODEL VARIABLES
# =============================================================================
# We apply listwise deletion only for the variables used in at least one
# model specification. Variables used only in robustness checks (nationality)
# are excluded from the main deletion step to preserve sample size.

core_vars <- c("log_hourly_wage", "gender", "education", "tenure",
               "age_group", "occupation", "sector", "contract_type",
               "working_day", "region", "firm_size", "firm_sector",
               "supervisory_role", "market_scope", "collective_bargaining")

df_main <- df_factors %>%
  drop_na(all_of(core_vars))

cat(sprintf("\nMain analytical dataset (Spec 1 & 2): n = %d\n", nrow(df_main)))
cat(sprintf("  Dropped for missing core vars: %d (%.2f%%)\n",
    nrow(df_filtered) - nrow(df_main),
    (nrow(df_filtered) - nrow(df_main)) / nrow(df_filtered) * 100))

# Full dataset for Spec 3 (DML) — also requires nationality
df_dml <- df_main %>%
  drop_na(nationality)

cat(sprintf("DML dataset (Spec 3, incl. nationality): n = %d\n", nrow(df_dml)))
print(core_vars)
head(df_main[,56:77])
df_main %>% 
  select(gender:last_col()) %>% 
  head()
glimpse(df_main[, 57:77])
df_main[,57:77] %>% head() %>% print(width = Inf) #Fig 3.6 
head(df_dml)
```

```{r}
# =============================================================================
# SECTION 8: THE THREE MODEL SPECIFICATIONS
# =============================================================================
#
# The three specifications are nested — each is a strict superset of the
# previous. This nesting is intentional: the progression of the gender
# coefficient from Spec 1 to Spec 3 is itself a result, showing how
# institutional variables (MERCADO, REGULACION) affect the estimate.
#
# References for specification choices:
#   Mincer (1974)          — core human capital equation
#   Goldin (2014)          — hours/flexibility mechanism
#   Blau & Kahn (2017)     — comprehensive controls benchmark
#   Chernozhukov et al.    — DML high-dimensional controls
#   Card, Cardoso &        — collective bargaining
#     Kline (2016)
#   Simón et al. (2008)    — Spanish institutional controls

# ── Spec 1: OLS Baseline ─────────────────────────────────────────────────────
# The canonical Mincer wage equation augmented with job and firm characteristics.
# Every variable here has a direct counterpart in the wage gap literature.
# This is also the specification used for the THREE models in Oaxaca-Blinder:
# (a) pooled, (b) male-only, (c) female-only.
spec1_controls <- c(
  # Human capital (Mincer 1974, Goldin 2014)
  "education",        # ESTU:    returns to schooling
  "tenure",           # ANOANTI: firm-specific experience
  "tenure_sq",        # ANOANTI²: concave experience-wage profile
  "age_group",        # ANOS2:   general labour market experience proxy

  # Job characteristics (Blau & Kahn 2017)
  "occupation",       # CNO1:    within-firm hierarchy position
  "sector",           # CNACE:   industry wage premium
  "contract_type",    # TIPOCON: permanent vs. temporary contract penalty
  "working_day",      # TIPOJOR: full vs. part-time (beyond hourly adjustment)
  "supervisory_role", # RESPONSA: managerial responsibility premium

  # Firm characteristics
  "region",           # NUTS1:   local labour market conditions
  "firm_size",        # ESTRATO2: size-wage premium (Oi & Idson 1999)
  "firm_sector"       # CONTROL: public sector pay compression
)

spec1_formula <- as.formula(
  paste("log_hourly_wage ~ gender +", paste(spec1_controls, collapse = " + "))
)
cat("\nSpec 1 formula:\n"); print(spec1_formula)

# ── Spec 2: OLS Extended + Oaxaca-Blinder ────────────────────────────────────
# Adds two institutional variables available in the EES but absent from most
# other wage datasets. Their inclusion allows us to test whether the gender
# coefficient is driven by institutional sorting (women in local-market, 
# centrally-bargained firms) rather than within-firm discrimination.
# If the gender coefficient barely changes from Spec 1 → Spec 2, that is
# evidence that institutional sorting is NOT confounding the OLS estimate.
spec2_controls <- c(
  spec1_controls,

  # Firm market exposure (Blau & Kahn 2017, Zweimüller et al. 2008)
  "market_scope",       # MERCADO: local → global (exporter wage premium)

  # Collective bargaining level (Card, Cardoso & Kline 2016)
  "collective_bargaining" # REGULACION: centralised vs. decentralised CB
)

spec2_formula <- as.formula(
  paste("log_hourly_wage ~ gender +", paste(spec2_controls, collapse = " + "))
)
cat("\nSpec 2 formula:\n"); print(spec2_formula)

# ── Spec 3: DML (high-dimensional, maximum confounding absorption) ────────────
# Adds nationality control for the DML nuisance models only.
# In DML, the nuisance functions E[Y|X] and E[D|X] are fitted with Random
# Forest — so additional variables cost nothing in interpretability and
# help absorb immigrant-female overlap confounding.
# NATIONALITY IS NOT INTERPRETED AS A CAUSAL VARIABLE — it is purely a
# confounding absorber in the ML nuisance step.
spec3_controls <- c(
  spec2_controls,
  "nationality"  # TIPOPAIS: absorbs immigrant-female overlap in DML nuisance
)

# Note: spec3_formula is defined but not used for OLS — only for DML
spec3_confounders <- spec3_controls  # used in DoubleMLData$new() x_cols

cat("\nSpec 3 DML confounders (", length(spec3_confounders), "variables):\n")
cat(paste(" -", spec3_confounders, collapse = "\n"), "\n")
```

```{r}
# =============================================================================
# SECTION 9: EXPLICITLY EXCLUDED VARIABLES & REASONS
# =============================================================================
# This section is documentation only — nothing is coded here.
# The rationale for each exclusion is stated for the thesis methodology
# chapter and for tribunal questions.

excluded_variables <- tribble(
  ~Variable,   ~Description,                          ~Reason_for_Exclusion,
  "IRPFMES",   "Monthly income tax withholding",       "Endogenous: IRPF is a function of salary (outcome), not a predictor. Including it creates reverse causality bias by construction.",
  "COTIZA",    "Social security contributions",         "Endogenous: SS contributions are set as a % of salary. Same problem as IRPFMES.",
  "BASE",      "Social security contribution base",     "Endogenous: computed from salary. Inclusion = regressing Y on a transformation of Y.",
  "SALBASE",   "Base salary (October)",                 "Component of RETRINOIN: including salary components as predictors of total salary creates perfect multicollinearity by construction.",
  "EXTRAORM",  "Monthly extraordinary payment",         "Component of RETRINOIN: same issue as SALBASE.",
  "COMSAL",    "Salary supplements (October)",          "Component of RETRINOIN: same issue as SALBASE.",
  "COMSALTT",  "Shift-work supplements",               "Component of RETRINOIN: same issue.",
  "GEXTRA",    "Annual extraordinary bonuses",          "Component of RETRINOIN: same issue.",
  "PHEXTRA",   "Overtime payments",                    "Component of RETRINOIN: same issue.",
  "RETRIIN",   "IT-related annual salary",              "Selection: incapacity spells introduce a non-random selection mechanism. Including IT salary mixes two different labour market states.",
  "FIJODISM",  "Months worked (discontinuous workers)", "Selection: applies only to fijo-discontinuo workers. Including it for the full sample introduces sample selection bias.",
  "FIJODISD",  "Days worked (discontinuous workers)",   "Same as FIJODISM.",
  "FACTOTAL",  "Sampling expansion factor",             "Design weight: this is a survey weight, not a covariate. Regression uses analytical weights if needed, not as a variable.",
  "ORDENTRA",  "Worker order number",                   "Administrative identifier: no economic content.",
  "DRELABD",   "Labour relationship duration (days)",   "Collinear with tenure and contract_type. The additional variation is tiny and interpretation is unclear in a stock survey.",
  "VESPIN",    "In-kind earnings (IT-related)",         "Same selection issue as RETRIIN.",
  "VESPNOIN",  "In-kind earnings (non-IT)",             "Minor component, not standard in wage gap literature. Adding complexity without proportional identification gain.",
  "JSP1/JSP2", "Weekly pacted hours",                   "Redundant with JAP (annual pacted hours). We use JAP directly in the dependent variable construction.",
  "HEXTRA",    "Overtime hours",                        "Partly endogenous to sector/occupation. Used descriptively in EDA but not in regression models."
)

cat("\n=== EXCLUDED VARIABLES (", nrow(excluded_variables), "variables) ===\n")
print(excluded_variables %>% select(Variable, Reason_for_Exclusion))
```

```{r}
# =============================================================================
# SECTION 10: FINAL DATASETS & QUICK DIAGNOSTICS
# =============================================================================

cat("\n=== FINAL DATASETS ===\n")
cat(sprintf("df_main  (Spec 1 & 2, OLS + OB): n = %d | vars = %d\n",
    nrow(df_main), ncol(df_main)))
cat(sprintf("df_dml   (Spec 3, DML):           n = %d | vars = %d\n",
    nrow(df_dml), ncol(df_dml)))

# Raw gender wage gap (before any controls)
gap_summary <- df_main %>%
  group_by(gender) %>%
  summarise(
    n              = n(),
    mean_log_wage  = round(mean(log_hourly_wage, na.rm = TRUE), 4),
    median_hourly  = round(median(hourly_wage,   na.rm = TRUE), 2),
    pct_fulltime   = round(mean(working_day == "Full-time", na.rm = TRUE) * 100, 1),
    pct_university = round(mean(education >= "University Diploma", na.rm = TRUE) * 100, 1),
    mean_tenure    = round(mean(tenure, na.rm = TRUE), 1),
    .groups = "drop"
  )
cat("\nDescriptive statistics by gender:\n")
print(gap_summary)

# Raw log-wage gap
raw_gap <- gap_summary$mean_log_wage[gap_summary$gender == "Female"] -
           gap_summary$mean_log_wage[gap_summary$gender == "Male"]
cat(sprintf("\nRaw unadjusted log-wage gap (F - M): %.4f\n", raw_gap))
cat(sprintf("Implied %% gap:                       %.2f%%\n",
    (exp(raw_gap) - 1) * 100))

# Gender composition of key variables (spot-check for plausibility)
cat("\nGender split by occupation group (top 5 most common):\n")
df_main %>%
  count(occupation, gender) %>%
  pivot_wider(names_from = gender, values_from = n, values_fill = 0) %>%
  mutate(total = Male + Female,
         pct_female = round(Female / total * 100, 1)) %>%
  arrange(desc(total)) %>%
  slice_head(n = 5) %>%
  print()

cat("\nGender split by collective bargaining level:\n")
df_main %>%
  count(collective_bargaining, gender) %>%
  pivot_wider(names_from = gender, values_from = n, values_fill = 0) %>%
  mutate(pct_female = round(Female / (Male + Female) * 100, 1)) %>%
  print()

cat("\nMarket scope distribution:\n")
print(table(df_main$market_scope, useNA = "always"))

cat("\n=== PREPROCESSING COMPLETE ===\n")
cat("Objects ready for analysis:\n")
cat("  df_main      — use for OLS (Spec 1, Spec 2) and Oaxaca-Blinder\n")
cat("  df_dml       — use for DML (Spec 3)\n")
cat("  spec1_formula — OLS Spec 1 formula object\n")
cat("  spec2_formula — OLS Spec 2 / Oaxaca-Blinder formula object\n")
cat("  spec3_confounders — character vector of DML X confounders\n")
cat("\nNext: source('01_EDA.R')\n")
```
